paper figures

### 2024-12-19 paper_figures.Rmd
add data

---
title: "paper figures"
author: "Benedikt Obermayer"
date: "2024/11/12"
output: 
  html_document:
    toc: true
    df_print: paged
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache=TRUE, cache.lazy=FALSE, message=FALSE, warning=FALSE,dev=c('pdf'))
library(ggpubr)
library(stats)
library(dplyr)
library(ggplot2)
library(tidyverse)
library(caret)
library(scales)
library(cowplot)
library(ComplexHeatmap)
library(circlize)
library(ggrepel)
library(car)
library(glmnet)
library(pROC)
library(readxl)
library(variancePartition)
library(RColorBrewer)
library(gtools)
library(grid)
library(DescTools)
library(lme4)
```

@@

diagnosis_colors <- c('control'='gray20','HD'='gray20',
                      'subclass'='lightslategray','CVID'='coral1',
                      'ICOS'='#8DD3C7','Kabuki'='#BEBADA',
                      'IgG4def'='burlywood4','TNFRSF13B'='palegreen4',
                      'other'='#BC80BD','AID'='#FDB462',
                      'ATM'='#A65628','LIG4'='mediumpurple2',
                      'NIPBL'='#FCCDE5','BRCA1'='darkkhaki',
                      'NFKB2'='#FFED6F','PI3K'='#80B1D3')

@@

```{r FigS3D_1, fig.width=1.4 * fig_scale, fig.height=1.7 * fig_scale}
tmp <- HD.var %>%  
  dplyr::left_join(columns, by='col') %>%
  dplyr::mutate(variable=factor(gsub('log10_nreads','# reads',variable),
                                levels=rev(c('donor','cohort','batch','sequencing','# reads','Residuals')))) %>%
  dplyr::filter(variable!='Residuals', category %in% c('diversity','diversity_raw'), variable!='cohort')

pvals <- tmp %>%
  dplyr::group_by(variable) %>%
  do(wc=wilcox.test(variance ~ category, data=.)) %>%
  summarise(variable, p=wc$p.value)

@@

pvals <- tmp %>% 
  dplyr::group_by(variable) %>%
  do(wc=wilcox.test(variance ~ category, data=., exact=FALSE)) %>%
  summarise(variable, p=wc$p.value)

@@

pvals <- tmp %>%
  dplyr::group_by(variable) %>%
  do(wc=wilcox.test(variance ~ averaging, paired=TRUE, data=., exact=FALSE)) %>%
  summarise(variable, p=wc$p.value)

@@

@@

@@

@@

ggplot(pcadat.subclass,aes(x=PC1,y=PC2)) +
  geom_point(aes(color=diagnosis),size=1) +
  geom_line(aes(group=donor, color=diagnosis)) +
  geom_polygon(data = pcadat.subclass %>%
                 dplyr::mutate(status=ifelse(diagnosis=='subclass', 'subclass', as.character(status))) %>%
                 dplyr::group_by(status) %>%
                 dplyr::slice(chull(PC1, PC2)), alpha = 0.25, aes(fill=status)) +
  my_theme() +
  #geom_label_repel(data = pcadat.subclass %>% 
  #                   dplyr::filter(diagnosis!='control'),
  #                 aes(label=gsub('FB_','',donor)),size=2, segment.size=.25, label.padding = unit(.5,'mm'), force=20, show.legend=FALSE) +
  scale_fill_manual(values=diagnosis_colors, guide='none') +
  scale_color_manual(values=diagnosis_colors) +
  guides(color=guide_legend(override.aes = list(shape=15,size=2), ncol=3)) +
  theme(aspect.ratio=pca$sdev[2]**2/pca$sdev[1]**2,
        legend.position='bottom',
        legend.box='horizontal') +
  labs(x=paste0('PC 1 (',100*signif(pca$sdev[1]**2/sum(pca$sdev**2),3),'%)'),
       y=paste0('PC 2 (',100*signif(pca$sdev[2]**2/sum(pca$sdev**2),3),'%)'),
       color='',shape='',fill='')
```

@@

tmp <- FB.data.pooled %>%
  dplyr::filter(sample_donor %in% FB_train_donors, diagnosis %in% c('control','subclass')) %>%
  gather(feature, frac_clusters, subclass.cols) %>%
  dplyr::mutate(feature=factor(gsub('frac_clusters_','',feature),
                               levels=gsub('frac_clusters_','',subclass.cols)))
ggplot(tmp, aes(x=feature, y=frac_clusters, color=diagnosis)) +
  geom_boxplot(position=position_dodge(width=.8),width=.7,outlier.shape=NA) +
  geom_point(position=position_dodge(width=.8), size=.5) +
  geom_line(data=tmp %>% dplyr::filter(diagnosis=='control'),
            aes(group=donor), position=position_nudge(x=-.2), alpha=.25, size=.25) +
  geom_line(data=tmp %>% dplyr::filter(diagnosis=='subclass'),
            aes(group=donor), position=position_nudge(x=+.2), alpha=.25, size=.25) +
  scale_color_manual(values=diagnosis_colors) +
  my_theme() +
  theme(legend.position=c(.8,.8)) +
  labs(color='',x='', y='fraction clusters')
```

@@

```{r FigS4X2, fig.width=2.5 * fig_scale, fig.height=2 * fig_scale}
FB.data.pooled %>%
  dplyr::mutate(alpha_ratio_FACS=pct_IgA/(pct_IgA + pct_IgG)) %>%
  ggplot(aes(x=alpha_ratio_FACS, y=alpha_ratio_clusters, color=status)) + 
  geom_abline(slope=1, intercept=0, size=.5) +
  stat_cor(method = "pearson", aes(label=..r.label..), size=2.5, show.legend=FALSE, label.x=.75, label.y=c(.4, .33)) +
  geom_point() + 
  my_theme() +
  scale_color_manual(values=diagnosis_colors) +
  labs(x='FACS', y='SWIBRID', title='IgA / (IgA + IgG)')
```

@@

```{r Fig4H, fig.width=2.5 * fig_scale, fig.height=1.5 * fig_scale}
set.seed(1)
train.here <- train & human.meta[train,'diagnosis'] != 'subclass'
cv_model <- cv.glmnet(as.matrix(human.X[train.here,]), human.meta[train.here,'status']!='control', alpha = 0, family='binomial')
mod <- glmnet(as.matrix(human.X[train.here,]), human.meta[train.here,'status']!='control', alpha = 0, family = "binomial",
                lambda = cv_model$lambda.min)
pred_1 <- predict(mod, type = "response", newx=as.matrix(human.X[test,]))
ROC_1 <- roc(human.meta[test,'status']!='control', pred_1)

@@

class_df <- data.frame(class=plyr::revalue(as.character(cutree(dend, k = 4)),
                                           c('2'='cluster3',
                                             '3'='cluster1',
                                             '1'='cluster2',
                                             '4'='cluster2')),
                       name=row.names(FB.X))
class_colors <- c('cluster3'=diagnosis_colors[['control']],
                  'cluster1'=cat_cols[['structural']],
                  'cluster2'=cat_cols[['diversity']])

@@

```{r FigS4C, fig.width=3 * fig_scale, fig.height=3 * fig_scale}
tmp <- data.frame(abs(pca$rotation)) %>%
  tibble::rownames_to_column('col') %>%
  dplyr::left_join(columns, by='col') %>%
  dplyr::mutate(col=factor(col, levels=columns %>% 
                             dplyr::filter(col %in% selected.cols) %>% 
                             dplyr::pull(col)),
                col_name=gsub("_downsampled|_switch",'',col))

@@

ggplot(tmp,aes(x=PC1,y=PC2, color=col, shape=col)) +
  geom_point(size=1) +
  my_theme() +
  scale_color_manual(values=cols_shapes %>%
                       dplyr::pull(var_col,col),
                     labels=cols_shapes %>%
                       dplyr::pull(col_name)) +
  scale_shape_manual(values=cols_shapes %>%
                       dplyr::pull(var_ind,col),
                     labels=cols_shapes %>%
                       dplyr::pull(col_name)) +
  theme(legend.key.height=unit(3,'mm'),
        aspect.ratio=pca$sdev[2]**2/pca$sdev[1]**2,
        legend.spacing=unit(0,'mm'),
        legend.position='bottom',
        legend.box='horizontal') +
  labs(x='abs. PC 1', y='abs. PC 2',
       color='',shape='',fill='',alpha='')
```

@@

```{r FigS4D, fig.width=3 * fig_scale, fig.height=1.5 * fig_scale}
ggplot(pca_hm %>% tibble::rownames_to_column('name') %>%
         dplyr::filter(dataset!='validation') %>%
         dplyr::left_join(class_df, by='name'), 
       aes(x=PC1,y=PC2, color=class, shape=dataset)) +
  geom_point(size=1) +
  geom_line(aes(group=donor)) +
  my_theme() +
  scale_shape_manual(values=c('testing'=15,'training'=16,'validation'=18)) +
  scale_color_manual(values=class_colors) +
  theme(legend.key.height=unit(3,'mm'),
        aspect.ratio=pca$sdev[2]**2/pca$sdev[1]**2) +
  labs(x=paste0('PC 1 (',100*signif(pca$sdev[1]**2/sum(pca$sdev**2),3),'%)'),
       y=paste0('PC 2 (',100*signif(pca$sdev[2]**2/sum(pca$sdev**2),3),'%)'),
       color='class')
```

@@

```{r Fig4J, fig.width=4.5 * fig_scale, fig.height=1.7 * fig_scale}
features.here <- c('frac_clusters_SG2',
                   'mean_cluster_size_downsampled',
                   'alpha_ratio_clusters',
                   'n_homology_switch',
                   'frac_breaks_duplications')

@@

ggplot(tmp, aes(x=group,y=value,fill=group)) +
  geom_boxplot(outlier.shape=NA, lwd=.5) +
  geom_jitter(width=.1,height=0,size=.5) +
  stat_compare_means(comparisons=list(c('cluster1','cluster3'),
                                      c('cluster2','cluster3'),
                                      c('control','ATM'),
                                      c('control','BRCA1')),
                     method='wilcox.test',label = "p.signif", hide.ns=TRUE, size=2, vjust=.5) +
  facet_wrap(.~col,ncol=length(features.here),scales='free_y') +
  my_theme() +
  coord_cartesian(clip='off') +
  scale_fill_manual(values=c(diagnosis_colors,class_colors)) +
  theme(axis.text.x=element_text(angle=90,hjust=1,vjust=.5),
        legend.position='none') +
  scale_x_discrete(breaks=c('cluster1','cluster2','cluster3','control','ATM','BRCA1'),
                   labels=c('cluster1','cluster2','cluster3','control','ATM','BRCA1')) +
  labs(x='',y='', color='diagnosis')
```

@@

```{r clinical_features, fig.width=3 * fig_scale, fig.height=2.5 * fig_scale}
pvals <- list()
for (x in c("Tetanus",
            "Diphteria",
            "PnPS")) {
  for (f in selected.cols) {
    data <- FB.data.pooled[FB.samples,c(x,f)] %>%
      tidyr::drop_na()
    form <- as.formula(paste0(x,'=="negative" ~ ',f))
    res <- glm(form, family='binomial', data=data) %>% summary()
    pvals[[paste0(x,'_',f)]] <- data.frame(pval=coef(res)[2,4],
                                           vacc=x,
                                           col=f)
  }
}

pvals <- list()
for (x in c("splenomegaly",
            "lymphoproliferation",
            "lung")) {
  for (f in selected.cols) {
    data <- FB.data.pooled[FB.samples,c(x,f,'status')] %>%
      dplyr::filter(status!='control') %>%
      tidyr::drop_na()
    form <- as.formula(paste0(x,'!="" ~ ',f))
    res <- glm(form, family='binomial', data=data) %>% summary()
    pvals[[paste0(x,'_',f)]] <- data.frame(pval=coef(res)[2,4],
                                           vacc=x,
                                           col=f)
  }
}

@@

```{r aggregated_FB_heatmap, fig.width=5.5 * fig_scale, fig.height=3.5 * fig_scale}
FB.X.aggregated <-  FB.data.pooled[FB.samples, selected.cols] %>%
  dplyr::mutate_if(is.numeric, scale) %>%
  tibble::rownames_to_column('sample') %>%
  gather(col, value, -sample) %>%
  dplyr::left_join(columns, by='col') %>%
  dplyr::group_by(sample, category) %>%
  dplyr::summarise(value=mean(abs(value))) %>%
  dplyr::left_join(FB.meta %>% tibble::rownames_to_column('sample'), by='sample') %>%
  #dplyr::filter(diagnosis!='control') %>%
  dplyr::select(sample,category,value) %>%
  spread(category,value) %>%
  tibble::column_to_rownames('sample')

hm2 <- Heatmap(t(FB.X.aggregated),
               #cluster_columns=dend,
               col=colorRamp2(c(0, 1.5), c("lightyellow2","steelblue3")),
               cluster_rows=TRUE,
               show_row_names=TRUE,
               show_column_names=TRUE,
               show_row_dend=TRUE,
               top_annotation=HeatmapAnnotation(df=FB.meta[row.names(FB.X.aggregated),c('diagnosis','sex','age',
                                                                                        'donor','timepoint'),drop=FALSE] %>% 
                                                  tibble::remove_rownames() %>%
                                                  dplyr::mutate(name=gsub('t','',gsub('FB_','',paste0(donor,'_',timepoint)))) %>%
                                                  dplyr::left_join(class_df,by='name') %>%
                                                  dplyr::select(-c(donor,timepoint)) %>%
                                                  tibble::column_to_rownames('name'),
                                                show_annotation_name=TRUE,
                                                show_legend=TRUE,
                                                simple_anno_size=unit(2,'mm'),
                                                annotation_legend_param=list(title_gp=gpar(fontsize=7),
                                                                             labels_gp=gpar(fontsize=7),
                                                                             grid_height = unit(3, "mm"),
                                                                             grid_width = unit(3, "mm")),
                                                annotation_name_gp=gpar(fontsize=8),
                                                col=list('diagnosis'=diagnosis_colors,
                                                         'dataset'=c('training'='gray60','testing'='gray20',
                                                                     'Charite'='red3','QP'='forestgreen'),
                                                         'class'=class_colors,
                                                         'EUROclass'=c('smB+21norm'='gray20',
                                                                       'smB+21low'='orange3',
                                                                       'smB-21norm'='lightblue3',
                                                                       'smB-21low'='darkseagreen2'),
                                                         'complication'=c('c'='coral3',
                                                                          'io'='olivedrab4'),
                                                         'age'=colorRamp2(c(20,50,80),c("yellow2","orange2","coral3")),
                                                         'sex'=c('male'='mistyrose2',
                                                                 'female'='lightseagreen'))),
               column_names_gp=gpar(fontsize=4),
               row_names_gp =gpar(fontsize = 7),
               row_dend_width = unit(7, "mm"),
               heatmap_legend_param=list(labels_gp=gpar(fontsize=7),
                                         title_gp=gpar(fontsize=8),
                                         legend_height=unit(10,'mm')),
               row_title=NULL,
               name='category score')

@@

```{r aggregated_FB_boxplots, fig.width=4.5 * fig_scale, fig.height=1.7 * fig_scale}
tmp <- FB.X.aggregated %>%
  tibble::rownames_to_column('sample') %>%
  dplyr::left_join(FB.meta %>% tibble::rownames_to_column('sample'), by='sample') %>%
  dplyr::mutate(name=gsub('t','',gsub('FB_','',paste0(donor,'_',timepoint)))) %>%
  dplyr::left_join(class_df,by='name') %>%
  gather(col,value,c('diversity','isotypes','breaks','context','structural'))

ggplot(tmp, aes(x=class,y=value,fill=class)) +
  geom_boxplot(outlier.shape=NA, lwd=.5) +
  geom_jitter(width=.1,height=0,size=.5) +
  stat_compare_means(comparisons=list(c('cluster1','cluster3'),
                                      c('cluster2','cluster3'),
                                      c('cluster1','cluster2')),
                     method='wilcox.test',label = "p.signif", hide.ns=TRUE, size=2, vjust=.5) +
  facet_wrap(.~col,ncol=length(features.here),scales='free_y') +
  my_theme() +
  coord_cartesian(clip='off') +
  scale_fill_manual(values=c(diagnosis_colors,class_colors)) +
  theme(axis.text.x=element_text(angle=45,hjust=1,vjust=1),
        legend.position='none') +
  labs(x='',y='category score', color='diagnosis')
```

@@

```{r Fig4K, fig.width=4. * fig_scale, fig.height=1.7 * fig_scale}
features.here <- c('frac_clusters_SG1',
                   'n_untemplated_switch',
                   'frac_breaks_intra',
                   'mean_length')

>>>>

---
title: "paper figures"
author: "Benedikt Obermayer"
date: "2024/12/10"
output: 
  html_document:
    toc: true
    df_print: paged
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache=TRUE, cache.lazy=FALSE, message=FALSE, warning=FALSE,dev=c('pdf'))
library(ggpubr)
library(stats)
library(dplyr)
library(ggplot2)
library(tidyverse)
library(caret)
library(scales)
library(cowplot)
library(ComplexHeatmap)
library(circlize)
library(ggrepel)
library(car)
library(glmnet)
library(pROC)
library(readxl)
library(variancePartition)
library(RColorBrewer)
library(gtools)
library(grid)
#library(DescTools)
library(lme4)
```

@@

diagnosis_colors <- c('control'='gray20','HD'='gray20',
                      'subclass'='lightslategray','CVID'='coral1',
                      'ICOS'='#8DD3C7','Kabuki'='#BEBADA',
                      'IgG4def'='burlywood4','TNFRSF13B'='palegreen4',
                      'other'='#BC80BD','AID'='darkgoldenrod3',
                      'ATM'='#A65628','LIG4'='mediumpurple2',
                      'NIPBL'='#FCCDE5','BRCA1'='darkkhaki',
                      'NFKB2'='#FFED6F','PI3K'='#80B1D3')

@@

```{r FigS3D_1, fig.width=1.4 * fig_scale, fig.height=1.7 * fig_scale}
tmp <- HD.var %>%  
  dplyr::left_join(columns, by='col') %>%
  dplyr::mutate(variable=factor(gsub('log10_nreads','# reads',variable),
                                levels=rev(c('donor','batch','sequencing','# reads','Residuals')))) %>%
  dplyr::filter(variable!='Residuals', category %in% c('diversity','diversity_raw'))

pvals <- tmp %>%
  dplyr::group_by(variable) %>%
  dplyr::do(wc=wilcox.test(variance ~ category, data=.)) %>%
  dplyr::summarise(variable, p=wc$p.value)

@@

pvals <- tmp %>% 
  dplyr::group_by(variable) %>%
  dplyr::do(wc=wilcox.test(variance ~ category, data=., exact=FALSE)) %>%
  dplyr::summarise(variable, p=wc$p.value)

@@

pvals <- tmp %>%
  dplyr::group_by(variable) %>%
  dplyr::do(wc=wilcox.test(variance ~ averaging, paired=TRUE, data=., exact=FALSE)) %>%
  dplyr::summarise(variable, p=wc$p.value)

@@

(this PCA uses `r dim(vincendeau.X)[2]` features)

@@

(this PCA uses `r dim(mouse.X)[2]` and `r dim(mouse.X.noSg)[2]` features)

@@

(this PCA uses `r dim(CH12.X)[2]` features)

@@

ggplot(pcadat.subclass,aes(x=PC1,y=PC2)) +
  geom_point(aes(color=diagnosis),size=1) +
  geom_line(aes(group=donor, color=diagnosis)) +
  geom_polygon(data = pcadat.subclass %>%
                 dplyr::mutate(status=ifelse(diagnosis=='subclass', 'subclass', as.character(status))) %>%
                 dplyr::group_by(status) %>%
                 dplyr::slice(chull(PC1, PC2)), alpha = 0.25, aes(fill=status)) +
  my_theme() +
  #geom_label_repel(data = pcadat.subclass %>% 
  #                   dplyr::filter(diagnosis!='control'),
  #                 aes(label=gsub('FB_','',donor)),size=2, segment.size=.25, label.padding = unit(.5,'mm'), force=20, show.legend=FALSE) +
  scale_fill_manual(values=diagnosis_colors, guide='none') +
  scale_color_manual(values=diagnosis_colors) +
  guides(color=guide_legend(override.aes = list(shape=15,size=2), ncol=3)) +
  theme(aspect.ratio=pca$sdev[2]**2/pca$sdev[1]**2,
        legend.position='bottom',
        legend.key.spacing.y=unit(0,'mm'),
        legend.box='horizontal') +
  labs(x=paste0('PC 1 (',100*signif(pca$sdev[1]**2/sum(pca$sdev**2),3),'%)'),
       y=paste0('PC 2 (',100*signif(pca$sdev[2]**2/sum(pca$sdev**2),3),'%)'),
       color='',shape='',fill='')
```

(this PCA uses `r dim(FB.X.subclass)[2]` features)

@@

tmp <- FB.data.pooled %>%
  dplyr::filter(sample_donor %in% FB_train_donors, diagnosis %in% c('control','subclass')) %>%
  gather(feature, frac_clusters, subclass.cols) %>%
  dplyr::mutate(feature=factor(gsub('frac_clusters_','',feature),
                               levels=gsub('frac_clusters_','',subclass.cols)))

pvals <- tmp %>%
  dplyr::group_by(feature) %>%
  dplyr::do(wc=wilcox.test(frac_clusters ~ diagnosis, data=.),
            y=max(.$frac_clusters)) %>%
  dplyr::summarise(feature, p.value=wc$p.value, y)

ggplot(tmp, aes(x=feature, y=frac_clusters)) +
  geom_boxplot(aes(color=diagnosis), position=position_dodge(width=.8),width=.7,outlier.shape=NA) +
  geom_point(aes(color=diagnosis), position=position_dodge(width=.8), size=.5) +
  geom_line(data=tmp %>% dplyr::filter(diagnosis=='control'),
            aes(group=donor), position=position_nudge(x=-.2), alpha=.25, size=.25) +
  geom_line(data=tmp %>% dplyr::filter(diagnosis=='subclass'),
            aes(group=donor), position=position_nudge(x=+.2), alpha=.25, size=.25) +
  geom_text(data=pvals, aes(label=stars.pval(p.value), y=1.1*y)) +
  scale_color_manual(values=diagnosis_colors) +
  my_theme() +
  coord_cartesian(clip='off') +
  theme(legend.position=c(.8,.8)) +
  labs(color='',x='', y='fraction clusters')
```

@@

```{r FigS4X2, fig.width=2.25 * fig_scale, fig.height=2 * fig_scale}
tmp <- FB.data.pooled %>%
  dplyr::mutate(alpha_ratio_FACS=pct_IgA/(pct_IgA + pct_IgG)) %>%
  dplyr::group_by(status) %>%
  dplyr::mutate(status_R=paste0(status," (R=",signif(cor(alpha_ratio_FACS,alpha_ratio_clusters,use='na.or.complete'),2),")"))
ggplot(tmp, aes(x=alpha_ratio_FACS, y=alpha_ratio_clusters, color=status)) + 
  geom_abline(slope=1, intercept=0, size=.5) +
  #stat_cor(method = "pearson", aes(label=..r.label..), size=2.5, show.legend=FALSE, label.x=.75, label.y=c(.4, .33)) +
  geom_point() + 
  my_theme() +
  scale_color_manual(values=diagnosis_colors, breaks=unique(tmp$status), labels=unique(tmp$status_R)) +
  labs(x='FACS', y='SWIBRID', title='IgA / (IgA + IgG)') +
  theme(legend.position=c(.8,.15),
        legend.box=element_blank())
```

@@

```{r Fig4H, fig.width=2. * fig_scale, fig.height=1.5 * fig_scale}
set.seed(1)
train.here <- train & human.meta[train,'diagnosis'] != 'subclass'
cv_model <- cv.glmnet(as.matrix(human.X[train.here,]), human.meta[train.here,'status']!='control', alpha = 0, family='binomial')
mod <- glmnet(as.matrix(human.X[train.here,]), human.meta[train.here,'status']!='control', alpha = 0, family = "binomial",
                lambda = cv_model$lambda.min)
pred_1 <- predict(mod, type = "response", newx=as.matrix(human.X[test,]))
ROC_1 <- roc(human.meta[test,'status']!='control', pred_1)

@@

class_df <- data.frame(class=plyr::revalue(as.character(cutree(dend, k = 4)),
                                           c('2'='cluster3',
                                             '3'='cluster1',
                                             '1'='cluster2',
                                             '4'='cluster2')),
                       name=row.names(FB.X))

class_colors <- c('cluster3'=diagnosis_colors[['control']],
                  'cluster1'=cat_cols[['structural']],
                  'cluster2'=cat_cols[['diversity']])

@@

```{r FigS4C, fig.width=3 * fig_scale, fig.height=4 * fig_scale}
tmp <- data.frame(abs(pca$rotation)) %>%
  tibble::rownames_to_column('col') %>%
  dplyr::left_join(columns, by='col') %>%
  dplyr::mutate(col=factor(col, levels=columns %>% 
                             dplyr::filter(col %in% selected.cols) %>% 
                             dplyr::pull(col)),
                col_name=gsub("_downsampled|_switch",'',col))

@@

ggplot(tmp,aes(x=PC1,y=PC2, color=col, shape=col)) +
  geom_point(size=1) +
  my_theme() +
  scale_color_manual(values=cols_shapes %>%
                       dplyr::pull(var_col,col),
                     labels=cols_shapes %>%
                       dplyr::pull(col_name)) +
  scale_shape_manual(values=cols_shapes %>%
                       dplyr::pull(var_ind,col),
                     labels=cols_shapes %>%
                       dplyr::pull(col_name)) +
  theme(legend.key.height=unit(3,'mm'),
        aspect.ratio=pca$sdev[2]**2/pca$sdev[1]**2,
        legend.spacing=unit(0,'mm'),
        legend.key.spacing.y=unit(0,'mm'),
        legend.position='bottom',
        legend.box='horizontal') +
  guides(color=guide_legend(ncol=2),
         shape=guide_legend(ncol=2)) +
  labs(x='abs. PC 1', y='abs. PC 2',
       color='',shape='',fill='',alpha='')
```

@@

```{r FigS4D, fig.width=2.5 * fig_scale, fig.height=2 * fig_scale}
ggplot(pca_hm %>% tibble::rownames_to_column('name') %>%
         dplyr::filter(dataset!='validation') %>%
         dplyr::left_join(class_df, by='name'), 
       aes(x=PC1,y=PC2, color=class, shape=dataset)) +
  geom_point(size=1) +
  geom_point(data=pcadat.combined %>% dplyr::filter(dataset=='validation'),
             aes(color=diagnosis), size=1) +
  geom_line(aes(group=donor)) +
  my_theme() +
  scale_shape_manual(values=c('testing'=16,'training'=16,'validation'=15), guide='none') +
  scale_color_manual(values=c(class_colors,diagnosis_colors), 
                     breaks=c(names(class_colors),
                              names(diagnosis_colors))) +
  guides(color=guide_legend(override.aes = list(shape=c(rep(16,3),rep(15,7)),size=2), ncol=4)) +
  theme(legend.key.height=unit(3,'mm'),
        legend.position='bottom',
        legend.key.spacing.y=unit(0,'mm'),
        aspect.ratio=pca$sdev[2]**2/pca$sdev[1]**2) +
  labs(x=paste0('PC 1 (',100*signif(pca$sdev[1]**2/sum(pca$sdev**2),3),'%)'),
       y=paste0('PC 2 (',100*signif(pca$sdev[2]**2/sum(pca$sdev**2),3),'%)'),
       color='')
```

@@

```{r Fig4J, fig.width=5.25 * fig_scale, fig.height=1.7 * fig_scale}
features.here <- c('frac_clusters_SG1',
                   'frac_clusters_SG2',
                   'mean_cluster_size_downsampled',
                   'n_homology_switch',
                   'n_untemplated_switch',
                   'frac_breaks_intra',
                   'frac_breaks_duplications')

@@

@@

```{r FigS4X4, fig.width=3 * fig_scale, fig.height=2.5 * fig_scale}
pvals <- list()
for (x in c("Tetanus",
            "Diphteria",
            "PnPS")) {
  for (f in selected.cols) {
    data <- FB.data.pooled[FB.samples,c(x,f)] %>%
      tidyr::drop_na()
    form <- as.formula(paste0(x,'=="negative" ~ ',f))
    res <- glm(form, family='binomial', data=data) %>% summary()
    pvals[[paste0(x,'_',f)]] <- data.frame(pval=coef(res)[2,4],
                                           vacc=x,
                                           col=f)
  }
}

for (x in c("splenomegaly",
            "lymphoproliferation",
            "lung")) {
  for (f in selected.cols) {
    data <- FB.data.pooled[FB.samples,c(x,f,'status')] %>%
      dplyr::filter(status!='control') %>%
      tidyr::drop_na()
    form <- as.formula(paste0(x,'!="" ~ ',f))
    res <- glm(form, family='binomial', data=data) %>% summary()
    pvals[[paste0(x,'_',f)]] <- data.frame(pval=coef(res)[2,4],
                                           vacc=x,
                                           col=f)
  }
}

@@

```{r FigS4X5, fig.width=5.5 * fig_scale, fig.height=3.5 * fig_scale}
FB.X.aggregated <-  FB.data.pooled[FB.samples, selected.cols] %>%
  dplyr::mutate_if(is.numeric, scale) %>%
  tibble::rownames_to_column('sample') %>%
  gather(col, value, -sample) %>%
  dplyr::left_join(columns, by='col') %>%
  dplyr::group_by(sample, category) %>%
  dplyr::summarise(value=mean(abs(value))) %>%
  dplyr::left_join(FB.meta %>% tibble::rownames_to_column('sample'), by='sample') %>%
  #dplyr::filter(diagnosis!='control') %>%
  dplyr::group_by(category) %>%
  dplyr::mutate(#value=(value-mean(value[diagnosis=='control'])/sd(value[diagnosis=='control'])),
                name=gsub('t','',gsub('FB_','',paste0(donor,'_',timepoint)))) %>%
  dplyr::select(name,category,value) %>%
  spread(category,value) %>%
  tibble::column_to_rownames('name')

FB.meta.aggregated <- FB.meta[,c('diagnosis','sex','age',
                                 'donor','timepoint'),drop=FALSE] %>% 
  dplyr::filter(!(diagnosis %in% c('control','subclass'))) %>%
  tibble::remove_rownames() %>%
  dplyr::mutate(name=gsub('t','',gsub('FB_','',paste0(donor,'_',timepoint)))) %>%
  dplyr::left_join(class_df,by='name') %>%
  dplyr::select(-c(donor,timepoint)) %>%
  tibble::column_to_rownames('name')

samples <- row.names(FB.meta.aggregated)
dend <- hclust(dist(FB.X.aggregated[samples,]))
dend <- reorder(as.dendrogram(dend),wts=order(match(FB.meta.aggregated %>%
                                                      tibble::rownames_to_column('sample') %>%
                                                      dplyr::arrange(class) %>%
                                                      dplyr::pull(sample),
                                                    samples)),agglo.FUN=mean)

hm2 <- Heatmap(t(FB.X.aggregated[samples,]),
               cluster_columns=dend,
               col=colorRamp2(c(0, 2), c("lightyellow2","darkorchid3")),
               cluster_rows=TRUE,
               show_row_names=TRUE,
               show_column_names=TRUE,
               show_row_dend=TRUE,
               top_annotation=HeatmapAnnotation(df=FB.meta.aggregated,
                                                show_annotation_name=TRUE,
                                                show_legend=TRUE,
                                                simple_anno_size=unit(2,'mm'),
                                                annotation_legend_param=list(title_gp=gpar(fontsize=7),
                                                                             labels_gp=gpar(fontsize=7),
                                                                             grid_height = unit(3, "mm"),
                                                                             grid_width = unit(3, "mm")),
                                                annotation_name_gp=gpar(fontsize=8),
                                                col=list('diagnosis'=diagnosis_colors,
                                                         'dataset'=c('training'='gray60','testing'='gray20',
                                                                     'Charite'='red3','QP'='forestgreen'),
                                                         'class'=class_colors,
                                                         'EUROclass'=c('smB+21norm'='gray20',
                                                                       'smB+21low'='orange3',
                                                                       'smB-21norm'='lightblue3',
                                                                       'smB-21low'='darkseagreen2'),
                                                         'complication'=c('c'='coral3',
                                                                          'io'='olivedrab4'),
                                                         'age'=colorRamp2(c(20,50,80),c("yellow2","orange2","coral3")),
                                                         'sex'=c('male'='mistyrose2',
                                                                 'female'='lightseagreen'))),
               column_names_gp=gpar(fontsize=4),
               row_names_gp =gpar(fontsize = 7),
               row_dend_width = unit(7, "mm"),
               heatmap_legend_param=list(labels_gp=gpar(fontsize=7),
                                         title_gp=gpar(fontsize=8),
                                         legend_height=unit(10,'mm')),
               row_title=NULL,
               name='category score')

@@

<!-- ```{r aggregated_FB_boxplots, fig.width=4.5 * fig_scale, fig.height=1.7 * fig_scale} -->
<!-- tmp <- FB.X.aggregated %>% -->
<!--   tibble::rownames_to_column('sample') %>% -->
<!--   dplyr::left_join(FB.meta %>% tibble::rownames_to_column('sample'), by='sample') %>% -->
<!--   dplyr::mutate(name=gsub('t','',gsub('FB_','',paste0(donor,'_',timepoint)))) %>% -->
<!--   dplyr::left_join(class_df,by='name') %>% -->
<!--   gather(col,value,c('diversity','isotypes','breaks','context','structural')) -->

<!-- ggplot(tmp, aes(x=class,y=value,fill=class)) + -->
<!--   geom_boxplot(outlier.shape=NA, lwd=.5) + -->
<!--   geom_jitter(width=.1,height=0,size=.5) + -->
<!--   stat_compare_means(comparisons=list(c('cluster1','cluster3'), -->
<!--                                       c('cluster2','cluster3'), -->
<!--                                       c('cluster1','cluster2')), -->
<!--                      method='wilcox.test',label = "p.signif", hide.ns=TRUE, size=2, vjust=.5) + -->
<!--   facet_wrap(.~col,ncol=length(features.here),scales='free_y') + -->
<!--   my_theme() + -->
<!--   coord_cartesian(clip='off') + -->
<!--   scale_fill_manual(values=c(diagnosis_colors,class_colors)) + -->
<!--   theme(axis.text.x=element_text(angle=45,hjust=1,vjust=1), -->
<!--         legend.position='none') + -->
<!--   labs(x='',y='category score', color='diagnosis') -->
<!-- ``` -->

@@
